Model Checking

### 2023-06-06 owen/Write Ups/03_existing_models.qmd
update the stan models to constrain scale params, even though forced to already by model
tidy up the current write up to include the fixes to the model, explore the relationship between posterior mean estimates and overall elo for players

```{r process data for this model}
#| eval: false

@@

bullet_60 <- lichess_data %>% 
  filter(Event == "Rated Bullet game") %>% 
  filter(TimeControl == "60+0")

@@

### this takes several hours, needs to be run on a cluster

stan_file <- "../Comp2.stan"

@@

```{r plot some parts of this model}
#| fig-height: 8

@@

fit_ppt <- readRDS(here("owen", "cluster_scripts", "Cluster_stan_ppt_2_n10.RDS"))

@@

@@

mcmc_hist(fit_ppt$draws(focal_gamma))
```

![Predictive Checks](../cluster_scripts/ppd_model2.png)

This model gives similar predictive performance to the previously proposed
model. However, there are some minor concerns with model fitting 
which should be addressed (convergence issues).

@@

-----

<!-- <!-- ## Model Comparison -->

<!-- We want to compare these potential models, which is very similar -->
<!-- to the model we fit previously. -->
<!-- The only change is that we now need the history for -->
<!-- both players, not just the focal player. -->

<!-- ```{r load in the initial data} -->
<!-- #| cache: true -->

<!-- lichess_data <- readRDS("../../rdata/lichess_pilot.RData") -->

<!-- bullet_60 <- lichess_data %>% -->
<!--   filter(Event == "Rated Bullet game") %>% -->
<!--   filter(TimeControl == "60+0") %>% -->
<!--   mutate(game_id = row_number()) -->

<!-- get_hist <- function(user, games, prev_n) { -->
<!--   hist_games <- games %>% -->
<!--     filter(White == user | Black == user) %>% -->
<!--     arrange(UTCDate, UTCTime) %>% -->
<!--     mutate(focal_white = ifelse(Username == White, 1, 0)) %>% -->
<!--     select(White:BlackElo, focal_white, game_id) %>% -->
<!--     mutate(focal_result = case_when( -->
<!--       (focal_white == 1 & Result == "1-0") ~ 1, -->
<!--       (focal_white == 0 & Result == "0-1") ~ 1, -->
<!--       (Result == "1/2-1/2") ~ 0.5, -->
<!--       .default = 0 -->
<!--       )) %>% -->
<!--     mutate(focal_win_prop = c(cumsum(focal_result[1:(prev_n - 1)])/(1:(prev_n -1)), -->
<!--                               roll_mean(focal_result, n = prev_n))) -->
<!--   hist_games -->
<!-- } -->

<!-- focal_players <- head(sort(table(c(bullet_60$Username)), decreasing = TRUE), 18) -->
<!-- ## if don't use all focal players here then removing data because they won't be -->
<!-- ## involved -->

<!-- top_players <- names(focal_players) -->

<!-- ## then get the opponents of these players -->

<!-- full_opponents <- bullet_60 %>% -->
<!--   filter(Username %in% top_players) %>% -->
<!--   select(White, Black) -->

<!-- all_opps <- c(full_opponents$White, full_opponents$Black) -->

<!-- opponents <- unique(all_opps) -->

<!-- ## then need to get the history of these players, but maybe need -->
<!-- ## to store game id -->
<!-- ## can just match using date and time I guess? -->
<!-- ## need to modify the function then to account for this -->

<!-- get_hist_opp <- function(user, games, prev_n) { -->
<!--   hist_games <- games %>% -->
<!--     filter(White == user | Black == user) %>% -->
<!--     arrange(UTCDate, UTCTime) %>% -->
<!--     mutate(opp_white = ifelse(user == White, 1, 0)) %>% -->
<!--     select(White:BlackElo, opp_white, game_id) %>% -->
<!--     mutate(opp_result = case_when( -->
<!--       (opp_white == 1 & Result == "1-0") ~ 1, -->
<!--       (opp_white == 0 & Result == "0-1") ~ 1, -->
<!--       (Result == "1/2-1/2") ~ 0.5, -->
<!--       .default = 0 -->
<!--       )) -->

<!--   ngames <- nrow(hist_games) -->

<!--   if(ngames < prev_n) { -->
<!--     hist_games <- hist_games %>% -->
<!--     mutate(opp_win_prop = opp_result/(1:ngames)) -->
<!--   } -->
<!--   else{ -->
<!--     hist_games <- hist_games %>% -->
<!--       mutate(opp_win_prop = c(cumsum(opp_result[1:(prev_n - 1)])/(1:(prev_n -1)), -->
<!--                               roll_mean(opp_result, n = prev_n))) -->
<!--   } -->
<!--   hist_games$player <- user -->
<!--   hist_games -->
<!-- } -->
<!-- ## so now the history is the history corresponding to the name -->
<!-- ## in the "player" column -->

<!-- remaining_opponents <- opponents[!(opponents %in% top_players)] -->

<!-- opp_hist <- map_dfr(remaining_opponents, get_hist_opp, bullet_60, prev_n = 5) %>% -->
<!--   as_tibble() -->

<!-- ## so this gives the history of the opponents for each of their -->
<!-- ## games, can use this to then first get their previous result, etc, -->
<!-- ## then join this with the other results -->

<!-- prev_opp <- opp_hist %>% -->
<!--   group_by(player) %>% -->
<!--   arrange(UTCDate, UTCTime) %>% -->
<!--   mutate(prev_opp_result = lag(opp_result, default = 0), -->
<!--          prev_opp_prop = lag(opp_win_prop, default = 0)) %>% -->
<!--   select(game_id, player, prev_opp_result, prev_opp_prop) -->

<!-- tidy_games <- map_dfr(top_players, get_hist, bullet_60, prev_n = 5) %>% -->
<!--   as_tibble() -->

<!-- all <- tidy_games %>% -->
<!--   left_join(prev_opp, by = "game_id") %>% -->
<!--   mutate(focal = ifelse(focal_white == 1, White, Black)) -->
<!-- ## this looks more reasonable now -->
<!-- ## this has the history of the opponent included, not the history -->
<!-- ## of the focal player yet -->

<!-- final_data <- all %>% -->
<!--   group_by(focal) %>% -->
<!--   arrange(UTCDate, UTCTime) %>% -->
<!--   mutate(prev_focal_result = lag(focal_result, default = 0), -->
<!--          prev_focal_prop = lag(focal_win_prop, default = 0)) %>% -->
<!--   ungroup() -->

<!-- ## then tidy this up for stan -->

<!-- ``` -->

<!-- This object `prev_game_both` should have the previous result of both players in -->
<!-- each game, if that game is in the dataset. -->
<!-- Similarly, `final_prev` has the games played by the focal players, along -->
<!-- with their previous result and their opponents previous result, if that -->
<!-- game was in the original dataset used (all 60 second bullet games here). -->

<!-- Then convert this into data for stan. -->

<!-- ```{r get stan data for simple model} -->

<!-- final_data -->

<!-- stan_prev <- final_data %>% -->
<!--   mutate(WhiteElo = as.numeric(WhiteElo), -->
<!--          BlackElo = as.numeric(BlackElo)) %>% -->
<!--   mutate(focal_user = ifelse(focal_white == 1, White, Black)) %>% -->
<!--   mutate(elo_diff = ifelse(focal_white == 1, -->
<!--                            WhiteElo - BlackElo, BlackElo - WhiteElo)) %>% -->
<!--   mutate(focal_id = match(focal_user, top_players)) %>% -->
<!--   select(focal_user, focal_id, focal_white, -->
<!--          prev_focal_result, prev_opp_result, focal_result) -->

<!-- stan_data <- list(N = nrow(stan_prev), -->
<!--                   J = length(top_players), -->
<!--                   y = stan_prev$focal_result, -->
<!--                   id = stan_prev$focal_id, -->
<!--                   focal_prev = stan_prev$prev_focal_result, -->
<!--                   opp_prev = stan_prev$prev_opp_result) -->

<!-- ``` -->

<!-- ```{r fit the simple stan model} -->
<!-- #| eval: false -->

<!-- ## remember no effect for non focal players in this at all -->
<!-- ## at the moment -->

<!-- stan_file <- "../Comp1.stan" -->

<!-- mod <- cmdstan_model(stan_file) -->

<!-- fit1_simple <- mod$sample(data = stan_data, -->
<!--                   seed = 123, -->
<!--                   chains = 4, -->
<!--                   parallel_chains = 4, -->
<!--                   refresh = 100) -->

<!-- ## this fits at least now... -->

<!-- fit1_simple$summary() -->
<!-- mcmc_hist(fit1_simple$draws(c("alpha", "beta"))) -->

<!-- ``` -->

<!-- ## Notes on these models -->

<!-- - How does the choice of focal player make a difference in these models? -->

<!-- ```{r thinking about the choice of focal} -->

<!-- table(bullet_60$Username) -->

<!-- tidy_bullet <- bullet_60 %>% -->
<!--   select(White:BlackElo, Username, game_id) %>% -->
<!--   mutate(opponent = ifelse(Username == White, Black, White), -->
<!--          focal_white = ifelse(Username == White, 1, 0)) -->

<!-- all_players <- c(tidy_bullet$Username, tidy_bullet$opponent) -->
<!-- players <- unique(all_players) -->

<!-- ## then just get all the history for this -->
<!-- all_hist <- map_dfr(players, get_hist_opp, bullet_60, prev_n = 5) %>% -->
<!--   as_tibble() %>% -->
<!--   select(game_id:player, UTCDate, UTCTime) -->

<!-- min_hist <- all_hist %>% select(game_id:player) -->

<!-- ## this should have each game appear twice, for each of the two players -->
<!-- ## involved, so need to figure out how to handle that best -->
<!-- ## left join with tidy games matching player to focal and then -->
<!-- ## to opponent, rename the columns as needed -->

<!-- tidy_stan <- tidy_bullet %>% -->
<!--   left_join(min_hist, by = c("game_id", "Username" = "player")) %>% -->
<!--   arrange(UTCDate, UTCTime) %>% -->
<!--   rename(focal_result = opp_result, focal_win_prop = opp_win_prop) %>% -->
<!--   left_join(min_hist, by = c("game_id", "opponent" = "player")) %>% -->
<!--   as_tibble() -->

<!-- ## this is all good, just need to lag both and add them in, -->
<!-- ## and also get the ids for the stan code -->

<!-- ## getting the overall current win probabilities for each player also, -->
<!-- ## up to and including the current game -->

<!-- player_avg <- all_hist %>% -->
<!--   group_by(player) %>% -->
<!--   arrange(UTCDate, UTCTime, .by_group = TRUE) %>% -->
<!--   mutate(nrow = row_number()) %>% -->
<!--   mutate(curr_avg = cumsum(opp_result)/nrow) %>% -->
<!--   select(game_id, player, curr_avg) -->

<!-- global_avg <- all_hist %>% -->
<!--   group_by(player) %>% -->
<!--   summarise(total_avg = mean(opp_result)) -->

<!-- ## then just join this in also to use it -->

<!-- ``` -->

<!-- ```{r getting the final data for potential stan model} -->

<!-- final_data <- tidy_stan %>% -->
<!--   left_join(player_avg, by = c("Username" = "player", "game_id")) %>% -->
<!--   rename(focal_avg = curr_avg) %>% -->
<!--   left_join(player_avg, by = c("opponent" = "player", "game_id")) %>% -->
<!--   rename(opp_avg = curr_avg) %>% -->
<!--   group_by(Username) %>% -->
<!--   arrange(UTCDate, UTCTime) %>% -->
<!--   mutate(focal_prev_result = lag(focal_result, default = 0), -->
<!--          focal_runn_avg = lag(focal_avg, n = 2, default = 0)) %>% -->
<!--   ungroup() %>% -->
<!--   group_by(opponent) %>% -->
<!--   arrange(UTCDate, UTCTime) %>% -->
<!--   mutate(opp_prev_result = lag(opp_result, default = 0), -->
<!--          opp_runn_avg = lag(opp_avg, n = 2, default = 0)) %>% -->
<!--   ungroup() %>% -->
<!--   mutate(focal_id = match(Username, players), -->
<!--          opp_id = match(opponent, players)) %>% -->
<!--   filter(focal_result != 0.5) %>% -->
<!--   select(-focal_avg, - opp_avg, - opp_result, -->
<!--          - focal_win_prop, - opp_win_prop) -->

<!-- final_data -->
<!-- ## remove some unused columns -->

<!-- norm_data <- final_data %>% -->
<!--   left_join(global_avg, by = c("Username" = "player")) %>% -->
<!--   rename(focal_global_avg = total_avg) %>% -->
<!--   left_join(global_avg, by = c("opponent" = "player")) %>% -->
<!--   rename(opp_global_avg = total_avg) %>% -->
<!--   mutate(focal_prev_result = focal_prev_result - focal_global_avg, -->
<!--          focal_runn_avg = focal_runn_avg - focal_global_avg, -->
<!--          opp_prev_result = opp_prev_result - opp_global_avg, -->
<!--          opp_runn_avg = opp_runn_avg - opp_global_avg) -->

<!-- ``` -->

<!-- Now this matches nicely, contains their historic winning average up to 2 games -->
<!-- previously, along with their result in the previous game. -->

<!-- Note that this has not been standardized currently. -->

<!-- ```{r} -->

<!-- norm_data -->

<!-- stan_data_all <- list(N = nrow(norm_data), -->
<!--                       J = length(players), -->
<!--                       y = norm_data$focal_result, -->
<!--                       focal_id = norm_data$focal_id, -->
<!--                       opp_id = norm_data$opp_id, -->
<!--                       focal_prev = norm_data$focal_prev_result, -->
<!--                       opp_prev = norm_data$opp_prev_result, -->
<!--                       focal_avg = norm_data$focal_runn_avg, -->
<!--                       opp_avg = norm_data$opp_runn_avg) -->

<!-- saveRDS(stan_data_all, file = "../cluster_scripts/stan_data.RDS") -->

<!-- stan_file <- "../Comp2.stan" -->

<!-- mod2 <- cmdstan_model(stan_file) -->

<!-- fit2 <- mod2$sample(data = stan_data_all, -->
<!--                     seed = 123, -->
<!--                     chains = 4, -->
<!--                     parallel_chains = 4, -->
<!--                     refresh = 100, -->
<!--                     iter_sampling = 500, -->
<!--                     iter_warmup = 100) -->

<!-- ``` -->

<!-- ## Fitting this to all data -->

<!-- Fit this model to all data on the cluster, just to see the results. -->
<!-- The traceplots and R-hat for this model indicates it doesn't work well, -->
<!-- does not converge. -->

<!-- ```{r} -->
<!-- stan_fit <- readRDS("../cluster_scripts/Cluster_stan.RDS") -->

<!-- mcmc_hist(stan_fit$draws(c("alpha[1]", "alpha[2]"))) -->

<!-- r_hat <- rhat(stan_fit) -->

<!-- summary(r_hat) -->

<!-- which(r_hat > 1.15) -->
<!-- ## just problems with mu and tau for convergence here, otherwise good? -->

<!-- ``` -->

<!-- ## Comparing models on smaller data -->

<!-- Is there a small subset of players who play each other often that we can -->
<!-- use in the dataset that can be used to compare these models? -->

<!-- ```{r} -->

<!-- bullet_60 %>% -->
<!--   filter(White %in% focal_players) %>% -->
<!--   filter(Black %in% focal_players) -->

<!-- ## no games between any of the players used to scrape the data -->
<!-- ## would maybe need to get more players then, to use this for a comparison -->

<!-- ``` -->

<!-- ```{r} -->
<!-- library(chessR) -->

<!-- chessdotcom_leaders <- chessdotcom_leaderboard(game_type = "live_bullet") -->

<!-- top_bullet <- chessdotcom_leaders %>% top_n(10, wt = -rank) %>% -->
<!--   select(player_id, username) -->

<!-- # chessdotcom_hikaru_recent <- get_raw_chessdotcom(usernames = "Hikaru", -->
<!-- #                                                  year_month = c(202201:202205)) -->

<!-- games_22 <- top_bullet$username %>% -->
<!--   map_dfr(~get_raw_chessdotcom(.x, year_month = c(202201:202203))) -->

<!-- top_games <- games_22 %>% -->
<!--   filter(White %in% top_bullet$username) %>% -->
<!--   filter(Black %in% top_bullet$username) -->

<!-- ## get 500 games here then for 1 month, probably enough to fit an initial model -->
<!-- ## most of these games bullet too, so could use that -->

<!-- top_games %>% -->
<!--   group_by(White, Black, TimeClass) %>% -->
<!--   tally() %>% -->
<!--   arrange(-n) -->

<!-- bullet_games <- top_games %>% -->
<!--   filter(TimeClass == "bullet") -->

<!-- bullet_top10 <- saveRDS(bullet_games, file = "chess_com_bullet_10.RDS") -->

<!-- ``` -->

>>>>

```{r process data for this model}

@@

bullet_60 <- lichess_data %>% 
  filter(Event == "Rated Bullet game") %>% 
  filter(TimeControl == "60+0") %>% 
  mutate(game_id = row_number())

```{r}
#| eval: false

@@

stan_data_all <- readRDS("../cluster_scripts/stan_data_ppt_n10.RDS")
### this takes several hours, needs to be run on a cluster

stan_file <- "../Comp3.stan"

@@

```{r plot some parts of this model}
#| fig-height: 5
#| fig-cap: Player effects for the 18 players considered by the previous model.

@@

fit_ppt <- readRDS(here("owen", "cluster_scripts",
                        "Cluster_stan_ppt_2_n10_June5.RDS"))

focal <- paste0("alpha[", c(1, 127, 1207, 1263, 1519, 1635, 
                           1789, 2084, 2139, 2441, 2503, 2561, 
                           2679, 2707, 2781, 2836, 3533, 3714), "]")

## most of these alphas centered around 0, 1/2 slightly positive and
## 1/2 slightly negative. these player effects

```{r}
#| fig-height: 5
#| fig-cap: Longer run (10 game) winner effects for those 18 players.

@@

## some of these betas not 0 really, this is more interesting maybe
## beta corresponds to the long run effect (prev 10 games)

```{r}
#| fig-height: 5
#| fig-cap: Effect of deviation in previous game from previous 10.

@@

mcmc_hist(fit_ppt$draws(focal_gamma))

## some of these slightly positive or negative, but not much
## this corresponds to previous game deviation from prev 10 games

```{r}
#| fig-height: 5
#| fig-cap: The global parameters for this model.

mcmc_hist(fit_ppt$draws(c("mu1", "mu2", "mu3",
                          "tau1", "tau2", "tau3")))

![Predictive Checks](../cluster_scripts/ppd_model2_June5.png)

This model gives similar predictive performance to the previously proposed
model.

@@

- How do these winner effects vary with the rating of players,
and how their rating changes over time?

```{r}

## note that rhat(fit_ppt) should work, just too big here

white <- bullet_60 %>% 
  select(game_id, White, WhiteElo) %>% 
  rename(Player = White, ELO = WhiteElo)

black <- bullet_60 %>% 
  select(game_id, Black, BlackElo) %>% 
  rename(Player = Black, ELO = BlackElo)

game_elo <- bind_rows(white, black)

avg_ELO <- game_elo %>% 
  mutate(ELO = as.numeric(ELO)) %>% 
  group_by(Player) %>% 
  summarise(mean_rating = mean(ELO))

## need to order these correctly, based on the indexing used for stan

player_info <- tibble(Player = curr_players) %>% 
  mutate(id = row_number())

avg_ELO <- avg_ELO %>% 
  left_join(player_info, by = "Player") %>% 
  drop_na()

## then get the posterior means for each of the parameters for this.

alpha_draws <- as_draws_df(fit_ppt$draws(c("alpha")))
beta_draws <- as_draws_df(fit_ppt$draws(c("beta")))
gamma_draws <- as_draws_df(fit_ppt$draws(c("gamma")))

alpha_mean <- apply(alpha_draws, 2, mean)
beta_mean <- apply(beta_draws, 2, mean)
gamma_mean <- apply(gamma_draws, 2, mean)
## then get the id's from this, based on the names
names(alpha_mean)

coef_info <- tibble(mean_alpha = alpha_mean, 
                    mean_beta = beta_mean,
                    mean_gamma = gamma_mean) %>% 
  mutate(id = row_number())

avg_ELO %>% 
  left_join(coef_info, by = "id") %>% 
  ggplot(aes(mean_alpha, mean_rating)) +
  geom_point() + 
  labs(x = "Posterior Mean for Alpha", y = "Mean ELO")

avg_ELO %>% 
  left_join(coef_info, by = "id") %>% 
  ggplot(aes(mean_beta, mean_rating)) +
  geom_point() + 
  labs(x = "Posterior Mean for Beta", y = "Mean ELO")

avg_ELO %>% 
  left_join(coef_info, by = "id") %>% 
  ggplot(aes(mean_gamma, mean_rating)) +
  geom_point() + 
  labs(x = "Posterior Mean for Gamma", y = "Mean ELO")

top_id <- c(1, 127, 1207, 1263, 1519, 1635, 
                           1789, 2084, 2139, 2441, 2503, 2561, 
                           2679, 2707, 2781, 2836, 3533, 3714)

avg_ELO %>% 
  left_join(coef_info, by = "id") %>% 
  mutate(top = ifelse(id %in% top_id, "TOP", "REST")) %>% 
  ggplot(aes(mean_alpha, mean_rating, colour = top)) +
  geom_point() + 
  labs(x = "Posterior Mean for Alpha", y = "Mean ELO")

avg_ELO %>% 
  left_join(coef_info, by = "id") %>% 
  mutate(top = ifelse(id %in% top_id, "TOP", "REST")) %>% 
  ggplot(aes(mean_beta, mean_rating, colour = top)) +
  geom_point() + 
  labs(x = "Posterior Mean for Beta", y = "Mean ELO")

avg_ELO %>% 
  left_join(coef_info, by = "id") %>% 
  mutate(top = ifelse(id %in% top_id, "TOP", "REST")) %>% 
  ggplot(aes(mean_gamma, mean_rating, colour = top)) +
  geom_point() + 
  labs(x = "Posterior Mean for Gamma", y = "Mean ELO")

## this just gives you the players who play few games all lumped together 
## basically, which makes sense
